Generalized dimensionality reduction scheme for global optimization problems \thanks{This study was supported by the Russian Science Foundation, project No.\,16-11-10150.}

### 2019-01-17 2019/WCGO/WCGO.tex
All the text except the Results

@@

\begin{document}
%
\title{ Generalized scheme for dimensionality reduction based on nested 
optimization and Peano curve for global optimization problems
\thanks{This study was supported by the Russian Science Foundation, project 
No.\,16-11-10150.}}
%
%\titlerunning{Abbreviated paper title}
% If the paper title is too long for the running head, you can set 
% an abbreviated paper title here
%
\author{Konstantin Barkalov \orcidID{0000-0001-5273-2471} \and
Ilia Lebedev\orcidID{0000-0002-8736-0652}}
%
\authorrunning{K. Barkalov, I. Lebedev}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
\institute{Lobachevsky State University of Nizhny Novgorod, Nizhny Novgorod, 
Russia \email{konstantin.barkalov@itmm.unn.ru}}
%
\maketitle              % typeset the header of the contribution
%
\begin{abstract}
The abstract should briefly summarize the contents of the paper in
150--250 words.

\keywords{Dimensionality reduction \and Global optimization \and 
Multiextremal functions.}
\end{abstract}
%
%
%
\section{Introduction}

\section{The core global search algorithm}

According to the algorithm, the first two trials are executed at the ends of 
the interval  $[a,b]$, i.e., $x^0=a,\;x^1=b$. The function values $z^0=\varphi
(x^0),\;z^1=\varphi(x^1)$  are computed and the number $k$ is set to 1. In 
order to select the point of a new trial $x^{k+1}, k\geq 1,$  it is necessary 
to perform the following steps.

\textbf{Step 1.} Renumber by subscripts (beginning from zero) the points $x^i,
\:0\leq i\leq k$, of the previous trials in increasing order, i.e.,
\begin{displaymath}
a=x_0<x_1<\ldots <x_{k}=b.
\end{displaymath} 
Juxtapose to the points $x_i,\:1\leq i\leq k$,  the function values $z_i=
\varphi(x_i),\:0\leq i\leq k$.

\textbf{Step 2.} Compute the maximum absolute value of the first divided 
differences 
\begin{displaymath}
\mu=\max_{1\leq i\leq k}\frac{\left|z_i-z_{i-1}\right|}{\Delta_i}
\end{displaymath}
where $\Delta_i = x_i-x_{i-1}$. If the above formula yields a zero value, 
assume that $\mu = 1$.

\textbf{Step 3.} For each interval $(x_{i-1},x_i),1\leq i\leq k$,  calculate 
the characteristic
\begin{displaymath}
R(i)=r\mu(x_i-x_{i-1})+\frac{(z_i-z_{i-1})^2}{r\mu(x_i-x_{i-1})}-2(z_i+z_{i-1}
),
\end{displaymath} 
where $r>1$ is a predefined parameter of the method.

@@

\textbf{Step 5.} Execute the new trial at the point 
\begin{displaymath}
x^{k+1}=\frac{1}{2}(x_{t-1}+x_t) - \frac{z_t-z_{t-1}}{2r\mu}.
\end{displaymath}

The algorithm terminates if the condition $\Delta_t<\epsilon$ is satisfied; 
here $t$ is from (\ref{MaxR}) and $\epsilon>0$ is the preset accuracy. For 
estimation of the global solution, values
\[
z_k^\ast=\min_{0\leq i \leq k}\varphi(x^i), \ x_k^\ast=\arg \min_{0\leq i \leq
 k}\varphi(x^i).
\]
are selected.

\section{Dimensionality reduction}
\subsection{Dimensionality reduction using Peano-type space-filling curves}

\subsection{Nested optimization schemes}

The nested optimization scheme of dimensionality reduction is based on the 
well-known relation (see, e.g., [])
\begin{equation}\label{nested}
\min_{y \in D}\varphi(y) = \min_{a_1\leq y_1 \leq b_1}\min_{a_2\leq y_2 \leq 
b_2}...\min_{a_N\leq y_N \leq b_N}\varphi(y),
\end{equation}
which allows replacing the solving of multidimensional problem (\ref{multi_
problem}) by solving a family of one-dimensional subproblems related to each 
other recursively.

@@

Then, according to relation (\ref{nested}), solving of multidimensional 
problem (\ref{multi_problem}) is reduced to solving a one-dimensional problem 
\begin{equation}\label{nested_1}
%\varphi^1(y_1^\ast) = 
\min_{a_1\leq y_1\leq b_1}\varphi^1(y_1).
\end{equation}
But in order to evaluate the function $\varphi^1$ at a fixed point $y_1$ it 
is necessary to solve the one-dimensional problem of the second level
\begin{equation}
\varphi^1(y_1) = \min_{a_2\leq y_2\leq b_2}\varphi^2(y_1,y_2),
\end{equation}
and so on up to the univariate minimization of the function $\varphi^N(y_1
,...,y_N)$ with fixed coordinates $y_1,...,y_{N-1}$ at the $N$-th level of 
recursion.

>>>>

%%%%ДОБАВИЛ ДЛЯ РУССКОГО ТЕКСТА
\usepackage[utf8x]{inputenc}
\usepackage[english,russian]{babel}
\usepackage{cmap}
%%%%

@@

\begin{document}
%
\title{Generalized dimensionality reduction scheme 
for global optimization problems
\thanks{This study was supported by the Russian Science Foundation, project 
No.\,16-11-10150.}}
%
\titlerunning{Generalized dimensionality reduction scheme}
% If the paper title is too long for the running head, you can set 
% an abbreviated paper title here
%
\author{Konstantin Barkalov %\orcidID{0000-0001-5273-2471}  
\and Ilia Lebedev %\orcidID{0000-0002-8736-0652}
}
%
\authorrunning{K. Barkalov, I. Lebedev}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
\institute{Lobachevsky State University of Nizhni Novgorod, Nizhni Novgorod, 
Russia \email{konstantin.barkalov@itmm.unn.ru}}
%
\maketitle              % typeset the header of the contribution
%
\begin{abstract}
The abstract should briefly summarize the contents of the paper in
150--250 words.

\keywords{Dimensionality reduction \and Global optimization \and 
Multiextremal functions.}
\end{abstract}
%
%
%
\section{Introduction}
This paper considers ''black-box'' global optimization problems of the following form:
\begin{eqnarray}\label{main_problem}
& \varphi(y^\ast)=\min{\left\{\varphi(y):y\in D\right\}},\\
& D=\left\{y\in R^N: a_i\leq y_i \leq b_i, 1\leq i \leq N\right\}. \nonumber
\end{eqnarray}
The objective function is assumed to satisfy the Lipschitz condition 
\[
\left|\varphi(y')-\varphi(y'')\right|\leq L\left\|y'-y''\right\|,\; y',y'' \in D,\; 0<L<\infty,
\]
with the constant $L$ unknown a priori.

\Russian

Известным способом решения многоэкстремальных задач являются мультистартовые схемы (см., например, [8–10]). В этих схемах в области поиска размещается некоторая сетка, точки которой использутся как стартовые поиска экстремума локальным методом, а затем выбирается наименьший из найденных экстремумов. Отдельной проблемой в рамках данного подхода является выбор стартовых точек, осуществляемый, как правило, на основе метода Монте-Карло \cite{Zhigljavsky2008}. Этот подход вполне применим для задач с небольшим числом локальных минимумов, имеющих широкие области притяжения, но для задач с существенной многоэкстремальностью его эффективность резко падает.

В настоящее время для решения задач глобальной оптимизации широко используются генетические алгоритмы (см., например, \cite{Yang2013}), которые так или иначе основаны на идеях случайного поиска. В силу простоты реализации и использования они получили большую популярность, однако по качеству работы (численной оценкой которого может служить число корректно решенных задач из некоторого тестового набора) они существенно уступают детерминированным алгоритмам \cite{Kvasov2018,Sergeyev2018}.

Если говорить о детерминированных методах глобальной оптимизации, то многие методы данного класса основаны на различных способах разбиения области поиска на систему подобластей и последующего выбора наиболее перспективной подобласти для размещения очередного испытания (вычисления целевой функции). Результаты в данном направлении представлены в работах \cite{Zilinskas2010,Paulavicius2016,Evtushenko2013,Jones2009,Sergeyev2015}.

Наконец, для разработки методов многомерной оптимизации широко применяется подход, связанный с редукцией многомерной задачи к эквивалентной одномерной или системе одномерных подзадач и последующим решением одномерных задач эффективными методами оптимизации функций одной переменной. Используются две такие схемы: редукция на основе кривых, заполняющих пространство (кривых Пеано, или разверток) \cite{Strongin2000,Sergeyev2013}, и схема рекурсивной вложенной оптимизации \cite{Strongin2000,Grishagin2001}. В статье \cite{Grishagin2016} предложена адаптивная схема редукции, обобщающая классическую схему рекурсивной оптимизации. Адаптивная схема существенно повышает эффективность оптимизации по сравнению с базовым прототипом \cite{Grishagin2016_1}. В данной работе предлагается обобщение адаптивной схемы редукции размерности, комбинирующее использование вложенной оптимизации и кривых Пеано. При таком подходе вложенные подзадачи в адаптивной схеме могут быть как одномерными, так и многомерными. В последнем случае для редукции размерности вложенных подзадач используются развертки.

\section{The global search algorithm}
\label{SectionCore}

As a core problem we consider a one-dimensional multiextremal optimization 
problem
\begin{equation}\label{uni_problem}
\varphi^\ast = \varphi(x^\ast)=\min{\left\{\varphi(x):x\in \left[0,1\right] 
\right\}}
\end{equation}
with objective function satisfying the Lipschitz condition.

Let us give the description of the global search algorithm (GSA) applied for 
solving the above problem (according to \cite{Strongin2000}). 
GSA involves constructing a sequence of points $x^i$, where the values of the 
objective function $z^i=\varphi(x^i)$ are calculated. Let us call the 
function value calculation process the \textit{trial}. 
According to the algorithm, the first two trials are executed at the ends of 
the interval  $[a,b]$, i.e., $x^0=a,\;x^1=b$. The function values $z^0=\varphi
(x^0),\;z^1=\varphi(x^1)$  are computed and the number $k$ is set to 1. In 
order to select the point of a new trial $x^{k+1}, k\geq 1,$  it is necessary 
to perform the following steps.

\textbf{Step 1.} Renumber by subscripts (beginning from zero) the points $x^i,
\:0\leq i\leq k$, of the previous trials in increasing order, i.e.,
\[
a=x_0<x_1<\ldots <x_{k}=b.
\] 
Juxtapose to the points $x_i, 0\leq i\leq k$,  the function values $z_i=
\varphi(x_i), 0\leq i\leq k$.

\textbf{Step 2.} Compute the maximum absolute value of the first divided 
differences 
\begin{equation}\label{mu}
\mu=\max_{1\leq i\leq k}\frac{\left|z_i-z_{i-1}\right|}{\Delta_i}
\end{equation}
where $\Delta_i = x_i-x_{i-1}$. If the above formula yields a zero value, 
assume that $\mu = 1$.

\textbf{Step 3.} For each interval $(x_{i-1},x_i),1\leq i\leq k$,  calculate 
the characteristic
\begin{equation}\label{R}
R(i)=r\mu\Delta_i+\frac{(z_i-z_{i-1})^2}{r\mu\Delta_i}-2(z_i+z_{i-1}),
\end{equation} 
where $r>1$ is a predefined parameter of the method.

@@

\textbf{Step 5.} Execute the new trial at the point 
\begin{equation}\label{xk1}
x^{k+1}=\frac{1}{2}(x_{t-1}+x_t) - \frac{z_t-z_{t-1}}{2r\mu}.
\end{equation}

The algorithm terminates if the condition $\Delta_t<\epsilon$ is satisfied; 
here $t$ is from (\ref{MaxR}) and $\epsilon>0$ is the preset accuracy. For 
estimation of the global solution values
\[
z_k^\ast=\min_{0\leq i \leq k}\varphi(x^i), \ x_k^\ast=\arg \min_{0\leq i \leq
 k}\varphi(x^i).
\]
are selected. 
The theory of algorithm convergence is presented in \cite{Strongin2000}.

\section{Dimensionality reduction}
\subsection{Dimensionality reduction using Peano-type space-filling curves}
\label{SectionPeano}

The use of Peano curve $y(x)$ 
\begin{equation}
\left\{y\in R^N: -2^{-1}\leq y_i \leq 2^{-1}, 1 \leq i \leq N\right\}=\left\{
y(x):0\leq x \leq 1 \right\}
\end{equation}
unambiguously mapping the interval of real axis $[0,1]$ onto a 
$N$-dimensional cube is the first of the dimension reduction methods considered.
Problems of numerical construction of Peano-type space-filling curves and the 
corresponding theory are considered in detail in \cite{Strongin2000}, 
\cite{Sergeyev2013}. Here we will note that a numerically 
constructed curve (\textit{evolvent}) is $2^{-m}$ accurate approximation of 
the theoretical Peano curve, where $m$ is an evolvent construction parameter.

By using this kind of mapping it is possible to reduce the multidimensional 
problem~(\ref{main_problem}) to a univariate problem
\begin{equation}
\varphi(y^\ast)=\varphi(y(x^\ast))=\min{\left\{\varphi(y(x)): x\in[0,1]
\right\}}.
\end{equation}
An important property of such mapping is preservation of boundedness of 
function relative differences (see \cite{Strongin2000,Sergeyev2013}). If the 
function $\varphi(y)$ in the domain $D$ satisfies the Lipschitz condition, 
then the function $\varphi(y(x))$ on the interval $[0,1]$ will satisfy a 
uniform H{\"o}lder condition
\begin{equation}\label{Holder}
\left|\varphi(y(x_1))-\varphi(y(x_2))\right|\leq H\left|x_1-x_2\right|^{1/N},
\end{equation}
where the H{\"o}lder constant $H$ is linked to the Lipschitz constant $L$ by 
the relation
\begin{equation}
H=2L\sqrt{N+3}.
\end{equation}
Condition (\ref{Holder}) allows adopting the algorithm for solving the 
one-dimensional problems presented in Section \ref{SectionCore} for solving the 
multidimensional problems reduced to the one-dimensional ones. For this, the 
lengths of intervals $\Delta_i$  involved into rules (\ref{mu}),(\ref{R}) of the
algorithm are substituted by the lengths 
\begin{equation}
\Delta_i=\left(x_i-x_{i-1}\right)^{1/N}
\end{equation}
and the following expression is introduced instead of formula (\ref{xk1}):
\begin{equation}
x^{k+1} = \frac{x_t+x_{t-1}}{2} - \mathrm{sign}(z_t-z_{t-1})\frac{1}{2r}
\left[\frac{\left|z_t-z_{t-1}\right|}{\mu}\right]^N.
\end{equation}

\subsection{Nested optimization scheme}

The nested optimization scheme of dimensionality reduction is based on the 
well-known relation (see, e.g., [])
\begin{equation}\label{nested}
\min_{y \in D}\varphi(y) = \min_{a_1\leq y_1 \leq b_1}\min_{a_2\leq y_2 \leq 
b_2}...\min_{a_N\leq y_N \leq b_N}\varphi(y),
\end{equation}
which allows replacing the solving of multidimensional problem 
(\ref{main_problem}) by solving a family of one-dimensional subproblems 
related to each other recursively.

@@

Then, according to relation (\ref{nested}), solving of multidimensional 
problem (\ref{main_problem}) is reduced to solving a one-dimensional problem 
\begin{equation}\label{nested_1}
%\varphi^1(y_1^\ast) = 
\min_{a_1\leq y_1\leq b_1}\varphi^1(y_1).
\end{equation}
But in order to evaluate the function $\varphi^1$ at a fixed point $y_1$ it 
is necessary to solve the one-dimensional problem of the second level
\begin{equation}
\varphi^1(y_1) = \min_{a_2\leq y_2\leq b_2}\varphi^2(y_1,y_2),
\end{equation}
and so on up to the univariate minimization of the function $\varphi^N(y_1
,...,y_N)$ with fixed coordinates $y_1,...,y_{N-1}$ at the $N$-th level of 
recursion.

For the nested scheme presented above, a generalization (\textit{block nested 
optimization scheme}), which combines the use of evolvents and the nested 
scheme has been elaborated in \cite{Barkalov2016}.

Let us consider vector $y$ as a vector of block variables
\begin{equation}
y=(y_1,y_2,...,y_N)=(u_1,u_2,...,u_M),
\end{equation}
where the $i$-th block variable $u_i$ is a vector of vector $y$ components, 
taken serially i.e. $u_1=(y_1,y_2,...,y_{N_1})$, $u_2=(y_{N_1+1},y_{N_1+2}
,...,y_{N_1+N_2})$,..., $u_M=(y_{N-N_M+1},y_{N-N_M+2},...,y_{N})$, where $N_1+
N_2+...+N_M=N$.

Using the new variables, main relation of the nested scheme (\ref{nested}) 
can be rewritten in the form 
\begin{equation}\label{block_nested}
\min_{y \in D}\varphi(y) = \min_{u_1\in D_1}\min_{u_2\in D_2}...\min_{u_M \in 
D_M}\varphi(y),
\end{equation}
where the subdomains $D_i, 1 \leq i \leq M$, are projections of initial 
search domain $D$ onto the subspaces corresponding to the variables $u_i, 1 
\leq i \leq M$.

The formulae defining the method of solving of problem (\ref{main_problem}) 
based on relation (\ref{block_nested}), in general, are the same to the ones 
of nested scheme (\ref{nested_N})--(\ref{nested_1}). It is only necessary to 
substitute the original variables $y_i, 1\leq i \leq N$,  by the block 
variables $u_i, 1\leq i \leq M$.

At that, the nested subproblems 
\begin{equation}\label{block_nested_i}
\varphi^i(u_1,...,u_i) = \min_{u^{i+1}\in D_{i+1}} \varphi_{i+1}(u_1,...,u_i,u
_{i+1}), 1\leq i\leq M-1.
\end{equation}
in the block scheme are the multidimensional ones. The dimension reduction 
method based on Peano curves can be applied for solving these ones. It is a 
principal difference from the initial nested scheme.

\subsection{Block adaptive optimization scheme}

\Russian
Решение возникающего множества подзадач вида (\ref{nested_i}) (для nested optimization scheme) или вида (\ref{block_nested_i}) (для block nested optimization scheme) может быть организовано различными способами. 
Очевидный способ (детально проработанный в \cite{Grishagin2001,Grishagin2015} для nested optimization scheme и в \cite{Barkalov2014,Barkalov2016} для block nested optimization scheme) основан на решении поздадач в соответствии с рекурсивным порядком их порождения. Однако здесь возникает потеря значительной части информации о objective function при решении многомерной задачи. Иным  подходом является адаптивная схема, в которой все подзадачи решаются одновременно, что позволяет более полно учитывать информацию о многомерной задаче и за счет этого ускорять процесс ее решения.

Для случая одномерных поздадач данный подход был реализован и теоретически обоснован в \cite{Grishagin2016,Grishagin2016_1,Grishagin2018}. Настоящая работа предлагает обобщение адаптивной схемы для случая многомерных подзадач.
Дадим краткое описание ее основных элементов.

Пусть вложенные подзадачи вида (\ref{block_nested_i}) решаются с помощью многомерного алгоритма глобального поиска, описанного в Section \ref{SectionPeano}. Тогда каждой подзадаче (\ref{block_nested_i}) можно присвоить некоторое числовое
значение, называемое характеристикой этой задачи. К качестве такой характеристики можно взять значение from (\ref{MaxR}), т.е. максимальную характеристику интервалов, сформированных в данной задаче. В соответствии с правилом вычисления характеристик (ссылка), чем выше значение характеристики, тем более перспективной является подзадача для продолжения поиска в ней глобального минимума исходной задачи (ссылка)  . Поэтому на каждой итерации выбирается подзадача с максимальной характеристикой для проведения в ней очередного испытания. Это испытание либо вычисляет значение целевой функции $\varphi(y)$ (если выбранная подзадача принадлежала уровню $j=M$), либо порождает новые подзадачи согласно (\ref{block_nested_i}) при $j\leq M-1$. В последнем случае новые порожденные задачи добавляются к текущему множеству задач, вычисляются их характеристики и процесс повторяется. Завершение процесса оптимизации происходит, когда в корневой задаче выполняется условие остановки алгоритма, решающего эту задачу.

\section{Results of numerical experiments}

One of the well-known approaches to investigating and comparing multiextremal 
optimization algorithms is based on the application of these methods for 
solving a set of test problems generated randomly.

The numerical comparison of the algorithms has been carried out using the Grishagin test problems [] and the GKLS test problem generator \cite{Gaviano2003}.

\section{Conclusion}
